---
title: "Simulation of drug treatment reversing disease-induced gene expression"
author: "Jitao David Zhang"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    df_print: paged
    theme: spacelab
    mathjax: default
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: inline
params:
  echo: yes
  relative: FALSE
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.height=6, fig.width=6)
library(ribiosUtils)
library(ribiosIO)
library(ribiosPlot)
library(openxlsx)
library(tidyverse)
library(ggplot2)
library(MASS) # for mvrnorm
library(dplyr)
library(tidyr)
library(ggrepel)
library(cowplot)
set.seed(1887)
theme_set(theme_light(base_size=16))
```

In this script, I show that omics data, for instance gene expression, can be used to inform drug discovery given the conditions below. 

* First, data are collected from human, for instance patients and healthy donors, as well as biological models used in the preclinical study, for instance animals treated with a drug candidate and without treatment, or disease-mimicking cell models treated with a drug candidate and without treatment. Thus effects of disease and drug treatment can be therefore estimated respectively.
* Second, the changes in patients should be ideally causally regulated by the disease.
* Third, the causality between disease and gene expression changes should be the same or similar in the disease model system used in the preclinical study.

Under these conditions, a negative correlation between the disease effect (compared with healthy control) and the drug effect (compared with placebo) would suggest the likely effectiveness of the drug treatment in patients. The strengths of the three conditions, i.e. expression changes, causality between the disease and the target gene's expression, and the conservation of causal structures between systems, apparently influence our confidence in the inference. The quantitative relationship between the strengths and the confidence remains to be tested.

# 1) Boxplot of three groups (healthy, disease, treatment)

Let us first examine what would happen when a drug treatment, which is known to be safe and effective, is tested directly in patients, i.e. not involving preclinical models.

We simulate 3 groups with 10 patients each. Means: healthy = 5, disease = 8, treatment = 6; standard deviation = 1.


```{r boxplot-sim, fig.height=4, fig.width=6}
n <- 10
means <- c(healthy = 5, disease = 8, treatment = 6)
sd <- 1

df_box <- bind_rows(
  lapply(names(means), function(g) {
    data.frame(group = g,
               value = rnorm(n = n, mean = means[[g]], sd = sd))
  })
) %>%
  mutate(group=factor(group, c("healthy", "disease", "treatment")))

pat_boxplot <- ggplot(df_box, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8, outliers = FALSE) +
  scale_fill_manual(values=c("darkgray", royalbluered(5)[5],
                             royalbluered(5)[1])) +
  geom_jitter(width=0.15) +
  theme(plot.subtitle = element_text(size=15),
        axis.title = element_text(size=15),
        axis.text.x = element_text(size=15)) +
  labs(title = "A partially working treatment",
       subtitle = "The gene is causally regulated by the disease",
       y = "Expression", x="Patient group")
print(pat_boxplot)
```

# 2) PCA of 30 individuals (10 healthy, 10 disease, 10 treatment)

We simulate multivariate normal features (e.g. 5000 genes). Disease shifts the data along the first principal axis by affecting the expression of 1% of all genes, and treatment is placed between healthy and disease so it lies between the two groups in PCA space.


```{r pca-sim, fig.height=4, fig.width=6}
set.seed(2025)
p <- 5000
n_group <- 10

# group centers: healthy = 0 vector; disease shift along first feature(s)
healthy_center <- rep(0, p)
disease_center <- c(rep(5, round(p/100)),
                        rep(0, p-round(p/100))) # shift along 1/10 of the features
# treatment halfway between
treatment_center <- (healthy_center *0.7 + disease_center*0.3)

# generate data with modest noise
sigma_noise <- 0.2


generate_group <- function(center, n) {
  m <- matrix(rnorm(n * p, mean = 0, sd = sigma_noise), nrow = n, ncol = p)
  sweep(m, 2, center, FUN = "+")
}


mat_healthy <- generate_group(healthy_center, n_group)
mat_disease <- generate_group(disease_center, n_group)
mat_treatment <- generate_group(treatment_center, n_group)


mat_all <- rbind(mat_healthy, mat_disease, mat_treatment)
rownames(mat_all) <- paste0(rep(c("Healthy","Disease","Treatment"), each = n_group), "_", 1:n_group)


# PCA
pca_res <- prcomp(mat_all, scale. = TRUE)
npc_df <- data.frame(pca_res$x[,1:2])
colnames(npc_df) <- c("PC1","PC2")
npc_df$group <- rep(c("healthy","disease","treatment"), each = n_group)
npc_df_healthy <- npc_df %>% filter(group=="healthy")
npc_df_healthyZero <- npc_df %>%
  mutate(PC1=(PC1 - mean(npc_df_healthy$PC1))/sd(npc_df_healthy$PC1),
         PC2=(PC2 - mean(npc_df_healthy$PC2))/sd(npc_df_healthy$PC2)) %>%
  mutate(group=factor(group, c("healthy", "disease", "treatment")))

# Plot
pca_plot <- ggplot(npc_df_healthyZero, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  scale_color_manual(values=c("black", royalbluered(5)[5],
                             royalbluered(5)[1])) +
  stat_ellipse(level = 0.68, linetype = "dashed") +
  ##geom_text_repel(aes(label = rownames(mat_all)), size = 4, max.overlaps = 12) +
  labs(title = "Simulated clinical observations",
       subtitle = "10 samples/group, 5000 features (1% diff. exp.)",
       x = "Principal component 1 (PC1)", y = "Principal component 2 (PC2)") +
  theme(plot.subtitle = element_text(size=15),
        axis.title = element_text(size=15),
        axis.text.x = element_text(size=15))
print(pca_plot)
```

# 3) Profiling disease's effect and drug's effect in patients and preclinical models, respectively 

```{r two_way_profiling, fig.height=4, fig.width=6}
n <- 10
means <- c(healthy = 5, disease = 8, placebo=9, treatment = 6.5)
sd <- 1

df_box <- bind_rows(
  lapply(names(means), function(g) {
    data.frame(group = g,
               value = rnorm(n = n, mean = means[[g]], sd = sd))
  })
) %>%
  mutate(group=factor(group, c("healthy", "disease", "placebo", "treatment"))) %>%
  mutate(setting=factor(ifelse(group %in% c("healthy", "disease"),
                        "clinical/\nreal-world", "preclinical/\ndisease models")))

two_comp_boxplot <- ggplot(df_box, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8, outliers = FALSE) +
  scale_fill_manual(values=c("darkgray", royalbluered(5)[5],
                             "#353535", royalbluered(5)[1])) +
  geom_jitter(width=0.15) +
  facet_wrap(~setting, scale="free_x") +
  theme(plot.subtitle = element_text(size=15),
        axis.title = element_text(size=15),
        axis.text.x = element_text(size=15, angle = 15,
                                   hjust = sqrt(5/6), vjust=sqrt(5/6))) +
  labs(title = "Simulation: two comparisons",
       subtitle = "The gene is causally regulated by the disease",
       y = "Expression", x="")
print(two_comp_boxplot)
```

# 4) Scatterplot of 20,000 points with correlation -0.65

Gaussian distribution

```{r scatter, fig.height=5, fig.width=5, message=FALSE}
N <- 20000
rho <- -0.65

# covariance matrix for bivariate normal
Sigma <- matrix(c(1, rho, rho, 1), nrow = 2)


# simulate bivariate normal
z <- mvrnorm(n = N, mu = c(0,0), Sigma = Sigma)

# assemble data frame
df_scatter <- data.frame(x = z[, 1], y = z[, 2])

# quick check of empirical correlation (Spearman/Kendall are better for heavy tails)
emp_cor <- with(df_scatter, cor(x, y, method = "pearson"))
emp_cor_spearman <- with(df_scatter, cor(x, y, method = "spearman"))

lfc_scatter <- ggplot(df_scatter, aes(x = x, y = y)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  labs(title = "Changes in two comparisons",
       subtitle = sprintf("Log2 fold change (log2FC) of %s genes",
                          N),
       x = "Change Disease-Healthy (log2FC)",
       y = "Change Drug-Control (log2FC)",) +
  scale_x_continuous(limits=c(-3, 3), oob=scales::censor) +
  scale_y_continuous(limits=c(-3, 3), oob=scales::censor) +
  theme(plot.subtitle = element_text(size=12),
        axis.title = element_text(size=16),
        axis.text.x = element_text(size=16)) +
  annotate("text", x=-0.8, 
           y=3, label=sprintf("Spearman cor.:%1.2f", emp_cor_spearman),
           hjust=0, size=5) +
  coord_fixed() + coord_equal()

lfc_scatter
```

Finally we put all plots in panels.

```{r panel, fig.height=9, fig.width=11}
plot_grid(pat_boxplot, pca_plot,
          two_comp_boxplot, lfc_scatter,
          nrow=2, ncol=2, align="hv", byrow=TRUE,
          labels = c('A', 'B', 'C', 'D'), label_size = 21)
```


# Disclaimers

The script was adapted from a template generated by ChatGPT. I did the final check.

# Session information

```{r bedaInfo}
bedaInfo()
```

```{r sessionInfo}
sessionInfo()
```

# Appendix
