---
title: "Mathematical modelling in pharmacology"
author: "Jitao david Zhang"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
fontsize: 10pt
output: 
  beamer_presentation:
    includes:
      in_header: mypreamble.tex
    toc: true
    colortheme: dolphin
    fig_height: 4
    fig_width: 4
    fonttheme: structurebold
    theme: CambridgeUS
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                     message=FALSE,
                     warnings=FALSE)
library(ggplot2)
ggplot2::theme_update(text=element_text(size=14))
ggplot2::update_geom_defaults("point", list(size = 2))
```

# Introduction to Pharmacology

## What is pharmacology

## Why mathematical modelling in pharmacology

Mathematical modelling links the experimental observations and the events taking place on the molecular level between the drug and protein recognition sites. It can be used to 

* Gain insights about the system;
* Predict the bahavior of the system;
* Explain or resolve apparent conflicts in observed data;
* Guide optimal experiment design;

## The Langmuir adsoprtion isotherm

## Dose-response data

\begincols
  \begincol{.4\textwidth}
  Function 1:
  $$
  y = 78(1-e^{-.76[\textrm{A}]^.75})-2
  $$
  Function 2:
  $$
  y = \frac{80\textrm{[A]}}{\textrm{[A]}+1}
  $$
  \endcol
  \begincol{.6\textwidth} 
```{r doseResponse, fig.height=3, fig.width=5}
dr1 <- function(x) 78*(1-exp(-.76*x^.75))-2
dr2 <- function(x) 80*x/(x+1)

dose <- 10^seq(-3,3)
dr1Data <- data.frame(Function="Func1", Input=dose, Output=dr1(dose))
dr2Data <- data.frame(Function="Func2", Input=dose, Output=dr2(dose))
drData <- rbind(dr1Data, dr2Data) 
drPlot <- ggplot(drData, aes(x=Input, y=Output)) +
  facet_wrap(~Function, scale="free_y") +
  scale_x_log10() +
  geom_point() +
  geom_line()
print(drPlot)
```
  One data set fit to two functions of the same general shape within the given domain.
  \endcol
\endcols

## An ubiquitous input-output model in biology


$$
\textrm{Output} = \frac{\alpha \cdot \textrm{[Input]}}{\beta \cdot \textrm{[Input]} + \gamma} + \delta
$$

* Visualization
* Non-linearity

## The classical model of receptor function

## The operational model of receptor function

## Two-state theory

## The ternary complex model and the extended ternary complex model

## The cubic ternary complex model

# A case study of computational enzymology

# Software

## COPASI

# Conclusions