---
title: "Introduction to Applied Mathematics and Informatics In Drug Discovery (iAMIDD)"
author: "Jitao David Zhang"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
fontsize: 10pt
output: 
  beamer_presentation:
    includes:
      in_header: mypreamble.tex
    toc: true
    colortheme: dolphin
    fig_height: 4
    fig_width: 4
    fonttheme: structurebold
    theme: CambridgeUS
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Introduction

## About me

Jitao _David_ Zhang, \myemail.

* Experience
    * Bachelor in Biology (2006)
    * Master in Bioinformatics (2008)
    * Ph.D. in Computational Biology and Bioinformatics (2011)
    * Working at Pharma Research and Early Development (pRED), Roche Innovation Center Basel, F. Hoffmann-La Roche, Ltd.
* Research interests
    * Modelling, interpretation, and integration of biological data
    * Developing novel computational and experimental approaches for disease understanding and drug discovery
    * Developing sofware to translate mathematical and computational concepts to (re)usable tools

\takehome{My personal opinions do not represent those of Roche}

## Some thoughts about the course series

* We try to understand key concepts, impacts, and challenges of applied mathematics and informatics approaches together.
* Case studies are discussed both in the context of drug discovery process and with a brief discussion of relevant mathematics and informatics concepts.
* Selected open-source software solutions are introduced along with the concepts.
* I am open to any questions, suggestions, and critics any time. Please share them directly with me at \myemail.

\takehome{Share your opinion with me to improve the course!}

## About you

Will you mind telling us ...

1. Who are you
2. What are the key steps in your educational and personal experience
3. Why are you joining this course series
4. What do you expect from this course series
5. Any questions you have about this course series

\takehome{Thank you \textit{all} for sharing}

# An overview on drug discovery and development

## Drug discovery is a long and expensive process

TODO: the avastin slide from Roche

## Phases of drug discovery

TA
  ~ Target assessment, should we invest to make a molecule modulating protein X?
  
Screening
  ~ Many molecules are evaluated and few, so-called *hits*, are prioritised.
  
LI
  ~ Lead identification, identify promising molecules chemists wish to work on

LO
  ~ Lead optimization, optimize pharmacological profiles of lead compounds
  
Phase I
  ~ Mainly for safety, mostly in healthy volunteers, but also patients possible
  
Phase II
  ~ Safety and efficacy, proof-of-concept
  
Phase III
  ~ Efficacy and efficacy, in large population

Phase IV
  ~ Follow-up study post approval

## DD is under constant (r)evolution

TODO: more

Since late 19th century, drug discovery experienced several waves of revolution

* The first wave, in early 20th century, was mainly driven by chemistry and microbiology
* The second wave, starting from ~1960s, was mainly driven by genetics and molecular biology 
* The third wave, starting from ~1980s, was mainly driven by technologies 
* The fourth wave, starting from 1990s, was driven by omics, increasing power of hardware and software to extract information from data
* Thr fifth wave, which we are experiencing, is characterized by large amount of heterogeneous data, novel biomarkers, and increasing awareness of the importance of data-driven decisions
* The next wave???

## Target assessment: which target to work on?

* Genetics and molecular biology revolutionlized drug discovery

\takehome{A good TA can lead to better understanding of biology}

## Screening: find a few hundred molecules to start!

\takehome{Tailored screening stategy increases likelihood of success}

## LI and LO phase: work up on the chosen few

\takehome{Molecular interaction and QSAR guides lead optimization}

## Preclinical safety: do animals tolerate the drug?

\takehome{To make a drug safe is no less important than making it work}

## PK/PD modelling and simulation: how much and how often?

\takehome{PK/PD connects exposure, efficacy, and time}

## Clinical trials: is the drug safe and efficacous?

* Experiment design
* Power analysis
* PTS (probability of technical success)
* ...
\takehome{Clinical trials can be tranquilizing for statisticians!}

## Biomarkers

\begincols
  \begincol{.5\textwidth}

This slide has two columns.

  \endcol
\begincol{.5\textwidth}

```{r}
set.seed(1887)
plot(function(x) 1/(x-0.5)^2+log(x), pch=21, xlim=c(1,6), xaxt="n",
     ylab="A simulated biomarker", xlab="Time",
     bg=sample(colors(), 10), cex=1.5)
```

  \endcol
\endcols

\takehome{Biomarkers are important!}

## Special aspects of DD compared with other industries

* Drug discovery is one of the most interdisciplinary enterprises of human beings
    * Pros for quantitative minds: mathematics is the language
    * Challenges for quantitative minds: openness and effective communication is critical
* We are *doing what patients need next*
    * We save lifes by making new drugs or introducing new diagonistic tools
* Drug discovery is subject to strong regulatory regulation
    * Especially in the development process, interactions with health authorities (HA) are essential.
* Ethic considerations are important
    * Refinement, replacement, and reduction (3R) of animal models
    * Accessibility of drugs for patients

# An overview on the course series

## The curriculum

* Computational biology and bioinformatics
    * Biological sequence analysis
    * Omics and next-generation sequencing
    * Bioinformatics approaches to understanding MoA of compounds
    * Protein sequence and structure
* Chemoinformatics and computer-aided drug design
    * Chemical structure representation and search
    * Ligand-based drug design
    * Structure-based drug design
* Mathematical modelling 
    * Principles and applications of modelling in pharmacology
    * Pharmacokinetics (PK) and pharmacodynamics (PD) modelling
    * Clinical pharmacology and pharmacokinetics
* From bench to bed and back
    * Emerging non-invasive markers: imaging and digital biomarker
    * Clinical trials
    * From real-world data to causal analysis and inference

## Administrivia

* Oral examination
* An in-depth course in computational biology, bioinformatics and biostatistics in preclinical drug discovery will be provisionally offered in the spring semester.

# Conclusions and perspectives

## Conclusions

* Drug discovery is a long and expensive process consisted of many phases
* Mathematics and informatics are critical to many aspects the drug-discovery process
* In *iAMIDD*, we will focus on selected topics where an quantatitive approach makes huge impact

## Next week: 

## References

## Technical notes

* The document is published under GPL-3 license under

