---
title: "Introduction to Applied Mathematics and Informatics In Drug Discovery (iAMIDD)"
author: "Jitao David Zhang"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
fontsize: 10pt
output: 
  beamer_presentation:
    includes:
      in_header: mypreamble.tex
    toc: true
    colortheme: dolphin
    fig_height: 4
    fig_width: 4
    fonttheme: structurebold
    theme: CambridgeUS
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Introduction

## About me

Jitao _David_ Zhang, \myemail.

* Experience
    * Bachelor in Biology (2006)
    * Master in Bioinformatics (2008)
    * Ph.D. in Computational Biology and Bioinformatics (2011)
    * Working at Pharma Research and Early Development (pRED), Roche Innovation Center Basel, F. Hoffmann-La Roche, Ltd.
* Research interests
    * Modelling, interpretation, and integration of biological data
    * Developing novel computational and experimental approaches for disease understanding and drug discovery
    * Developing sofware to translate mathematical and computational concepts to (re)usable tools

\takehome{My personal opinions do not represent those of Roche}

## Some thoughts about the course series

* We try to understand key concepts, impacts, and challenges of applied mathematics and informatics approaches together.
* Case studies are discussed both in the context of drug discovery process and with a brief discussion of relevant mathematics and informatics concepts.
* Selected open-source software solutions are introduced along with the concepts.
* I am open to any questions, suggestions, and critics any time. Please share them directly with me at \myemail.

\takehome{Share your opinion with me to improve the course!}

## About you

Will you mind telling us ...

1. Who are you
2. What are the key steps in your educational and personal experience
3. Why are you joining this course series
4. What do you expect from this course series
5. Any questions you have about this course series

\takehome{Thank you \textit{all} for sharing}

# An overview on drug discovery and development

## Drug discovery is a long and expensive process

TODO: the avastin slide from Roche

## Phases of drug discovery

TA
  ~ Target assessment, should we invest to make a molecule modulating protein X?
  
Screening
  ~ Many molecules are evaluated and few, so-called *hits*, are prioritised.
  
LI
  ~ Lead identification, identify promising molecules chemists wish to work on

LO
  ~ Lead optimization, optimize pharmacological profiles of lead compounds
  
Phase I
  ~ Mainly for safety, mostly in healthy volunteers (HVs), but also patients possible
  
Phase II
  ~ Safety and efficacy, proof-of-concept
  
Phase III
  ~ Efficacy and efficacy, in large population

Phase IV
  ~ Follow-up study post approval

## DD is under constant (r)evolution

TODO: more

Since late 19th century, drug discovery experienced several waves of revolution

* The first wave, in early 20th century, was mainly driven by chemistry and microbiology
* The second wave, starting from ~1960s, was mainly driven by genetics and molecular biology 
* The third wave, starting from ~1980s, was mainly driven by technologies 
* The fourth wave, starting from 1990s, was driven by omics, increasing power of hardware and software to extract information from data
* Thr fifth wave, which we are experiencing, is characterized by large amount of heterogeneous data, novel biomarkers, and increasing awareness of the importance of data-driven decisions
* The next wave???

## Target assessment: which target to work on?

* Sometimes known as target identification and validation
* Genetics and molecular biology revolutionlized drug discovery
* Phenotypic and genotypic approaches

\takehome{A good TA can lead to better understanding of biology}

## Screening: find a few hundred molecules to start!

\takehome{Tailored screening stategy increases likelihood of success}

## Lead Identification (LI)

* Selection of most promising chemical series

## Lead Optimization (LO)

* Selection of most promising candidate(s)

\takehome{Molecular interaction and QSAR guides lead optimization}

## Formulation and delivery

* Biopharmaceutical properties
* Formulation to optimize use, safety, and efficacy

## Preclinical safety and toxicology

* ADME studies
    * **A**bsorption
    * **D**istribution
    * **M**etabolism
    * **E**xcretion
* GLP (Good Laboratory Process) toxicology studies
    * Species selection based on relevance and sensitivity
    * Target expression
    * On-target and off-target effects
    * Dose selection for human studies with PK/PD modelling and simulation: how much and how often?
    * Inform risk management in humans
* Investigative toxicology studies

The three **R** (replace, reduce, and refine) principle applies.

\takehome{To make a drug safe is no less important than making it work}

## Clinical trials: is the drug safe and efficacous?

* Experiment design
* Power analysis
* PTS (probability of technical success)
* ...
\takehome{Clinical trials can be tranquilizing for statisticians!}

## Phase I clinical trials

* Safety and tolerability
    * Define a safe clinical dose range
* Human pharmacology
    * The right dose for the right patient
* Studies in a small number of subjects
    * HV vs. patients
* Drug-drug interaction studies

## Phase II clinical trials

* Therapeutic exploration
* Safety in the target population
* Proof-of-concept studies
* Use of biomarkers to stratify patients
* Dose selection
* Pharmacokinetics/pharmacodynamics (PK/PD) modelling

## Phase III clinical trials

* Therapeutic confirmation
    * Sometimes known as *pivoral* trials
    * Usually multinational studies with many more patients than previous phases
* Safety evaluation for post-marketing risk management 
* Regulatory filing

## Phase IV clinical trials

* Therapeutic use
    * Post-registration monitoring
    * Further exploration/confirmation of safety and efficacy profiles
    * Real-life experience of the medicine
    * Pharmacoeconomic evaluations
* Global expansion
* Novel indications, sometimes known as *repurposing*.
* Investigator-led studies
* Epidemiologic studies

## Model- and data-driven approaches drive current trends in clinical development 

** Maybe we should put this in the clinical part of the presentation? **

* Seamless development programs, for instance
    * Integrated phase I-II studies
    * Move from phase I to phase III directly
    * Adaptive clinical studies
    * Staggered enrollment and interim analysis
* Predictions based on preclinical models
* Use PK/PD modelling and simulation
* Use of biomarkers for patient stratification

## Biomarkers

\begincols
  \begincol{.5\textwidth}

This slide has two columns.

  \endcol
\begincol{.5\textwidth}

```{r}
set.seed(1887)
plot(function(x) 1/(x-0.5)^2+log(x), pch=21, xlim=c(1,6), xaxt="n",
     ylab="A simulated biomarker", xlab="Time",
     bg=sample(colors(), 10), cex=1.5)
```

  \endcol
\endcols

\takehome{Biomarkers are important!}

## Special aspects of DD compared with other industries

* Drug discovery is one of the most interdisciplinary enterprises of human beings
    * Pros for quantitative minds: mathematics is the language
    * Challenges for quantitative minds: openness and effective communication is critical
* Scientifically challenging
    * Much remains unknown about biology and chemistry
* We are *doing what patients need next*
    * We save lifes by making new drugs or introducing new diagonistic tools
* Strong regulatory regulation
    * Especially in the development process, interactions with health authorities (HA) are essential.
* Economic implications
* Ethic considerations are important
    * Refinement, replacement, and reduction (3R) of animal models
    * Accessibility of drugs for patients

## Approval times

Kaitin KI et al., Clin Pharmacol Ther. 2011; 89(2): 183-188

## Considerations to start a drug-discovery program

* Efficacy
* Safety
    * Thorough evaluation of benefit and risk
    * Risk management
* Marketing
    * Differenation
    * Evidence of efficacy
    * Long-temr development plans
* Timing
* Regulatory strategy
* Clinical use

## More factors for consideration

* Medical need
* Business and strategic plan
* Therapeutic modality
* Indication (e.g. high risk versus primary care)
* First-in-class vs. follow-on
* Regulatory plan
* Internal capabilities and expertise
* Availability and predictability of non-clinical models
* Planned population in early clinical studies
* etc.

## Therapeutic modalities

* Small molecules
* Peptides
* Antibodies
* Therapetic proteins
* Oligonucleotides
* Antibody-drug conjugates
* Vaccines
* Gene therapy

## Today's drug development work is for tomorrow

** Need to start with the end in mind **

* Use of models/surrogate markers for extrapolation
    * Non-clinical --> clinical
    * HVs --> Patients
    * Phase II --> Phase III
    * Phase III --> Phase IV
    * Current indication(s) --> Novel indication(s)
* Safety always - Hoping for efficacy
    * Need to optimize risk benefit
    * Understand risk tolerance
    * Alternative therapeutic options
* Characterize and understand variability in response

## Success rates by therapeutic classes

DiMasi JA et al. Clin PHarmacol Ther. 2010;87(3):272-277

## Success rates as experienced by AstraZeneca

Morgan et al. NRDD. 2018; 17(3):167-181

## Drug discovery is a long and expensive process

PhRMA, Drug Development Process

# An overview on the course series

## The curriculum

* Bioinformatics and computational biology
    * Biological sequence analysis
    * Omics and next-generation sequencing
    * Bioinformatics approaches to understanding MoA of compounds
    * Protein sequence and structure
* Chemoinformatics and computer-aided drug design
    * Chemical structure representation and search
    * Ligand-based drug design
    * Structure-based drug design
* Mathematical modelling 
    * Principles and applications of modelling in pharmacology
    * Pharmacokinetics (PK) and pharmacodynamics (PD) modelling
    * Clinical pharmacology and pharmacokinetics
* From bench to bed and back: statistics, machine learning, and causal inference
    * Emerging non-invasive markers: imaging and digital biomarker
    * Clinical trials
    * From real-world data to causal analysis and inference

## What To and Not To Expect

\begincols
  \begincol{.5\textwidth}
**What do expect**

* Introduction to many different aspects of applied mathematics and informatics in drug discovery
* Awareness of how applied mathematics and informatics helps transforming good ideas into a medicine
* Raise interest, questions, and explore options for professional development
* Resources if you want to learn more about specific topics

  \endcol
  \begincol{.5\textwidth}
**What not to expect**

* Emphasis on one specific descipline
* Detailed descriptions of each topic
* Prompt and accurate answers for all questions (we will try to find them out, though!)

  \endcol
\endcols

## Administrivia

* Oral examination
* An in-depth course in computational biology, bioinformatics and biostatistics in preclinical drug discovery will be provisionally offered in the spring semester.

# Conclusions and perspectives

## Conclusions

* Drug discovery is a long and expensive process consisted of many phases
* Mathematics and informatics are critical to many aspects the drug-discovery process
* In *iAMIDD*, we will focus on selected topics where an quantatitive approach makes huge impact

## Next week: 

## References

## Acknowledgment

* Dr. Ignacio Rodriguez

## Technical notes

* The document is published under GPL-3 license under

