---
title: "Introduction to Applied Mathematics and Informatics In Drug Discovery (iAMIDD)"
author: "Jitao David Zhang"
date: "13 März 2019"
output: 
  beamer_presentation:
    includes:
      in_header: mypreamble.tex
    colortheme: dolphin
    fig_height: 4
    fig_width: 4
    fonttheme: structurebold
    theme: CambridgeUS
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Before we start ...

# Outline of today's course

# Phases of drug discovery

TA
  ~ Target assessment, should we invest to make a molecule modulating protein X?
  
Screening
  ~ Many molecules are evaluated and few, so-called *hits*, are prioritised.
  
LI
  ~ Lead identification, identify promising molecules chemists wish to work on

LO
  ~ Lead optimization, optimize pharmacological profiles of lead compounds
  
Phase I
  ~ Mainly for safety, mostly in healthy volunteers, but also patients possible
  
Phase II
  ~ Safety and efficacy, proof-of-concept
  
Phase III
  ~ Efficacy and efficacy, in large population

Phase IV
  ~ Follow-up study post approval
  
# Target assessment

\takehome{A good TA can lead to better understanding of biology}

# Screening

\takehome{Tailored screening stategy increases likelihood of success}

# LI and LO phase

\takehome{Molecular interaction and QSAR guides lead optimization}

# Preclinical safety

\takehome{To make a drug safe is no less important than making it work}

# PK/PD modelling and simulation

\takehome{PK/PD connects exposure, efficacy, and time}

# Clinical trials

* Experiment design
* Power analysis
* PTS (probability of technical success)
* ...
\takehome{Clinical trials can be tranquilizing for statisticians!}

# Biomarkers

\begincols
  \begincol{.5\textwidth}

This slide has two columns.

  \endcol
\begincol{.5\textwidth}

```{r}
set.seed(1887)
plot(function(x) 1/(x-0.5)^2+log(x), pch=21, xlim=c(1,6), xaxt="n",
     ylab="A simulated biomarker", xlab="Time",
     bg=sample(colors(), 10), cex=1.5)
```

  \endcol
\endcols

\takehome{Biomarkers are important!}

# Conclusions

* Phases
* Mathematics and informatics are critical to the drug-discovery process

# References

# Technical notes

See more setting options on [https://bookdown.org/yihui/rmarkdown/beamer-presentation.html](https://bookdown.org/yihui/rmarkdown/beamer-presentation.html).

A comprehensive tutorial on using Markdown to make beamer presentations with `pandoc` can be found at [https://pandoc.org/MANUAL.html#pandocs-markdown](https://pandoc.org/MANUAL.html#pandocs-markdown).
