# Introduction {#intro}

## Drug discovery as a multiscale modelling problem

Drug discovery is an interdisciplinary enterprise. To develop a efficacious and safe drug, it requires the joint effort of scientists of many fields, in particular chemistry, biology, medicine, physics, statistics, mathematics, informatics,. 

A common trait underlying contributions of different fields to drug discovery is that they presume models, and generate data to either validate, reject, or amend and improve the models. Drug mechanism and safety is profiled at multiple scales - molecular, cellular, and systematic levels - before a candidate can be tested in human in clinical trials. The *multiscale modelling process*, a term that I borrow from the field of physics and material science, lies right in the heart of applied mathematics and informatics in drug discovery. It sets the framework in which mathematical concepts and tools are used to describe complex biological systems and how drugs assert effects on them, and in which computers are used to translate the mathematical objects into human interpretable results.

In light of the multiscale modelling process, this course series introduces selected applications of mathematics and informatics in drug discovery. In particular, we focus on four areas:

1. Molecular-level modelling, which models structures of molecules and interactions between molecules using molecular modelling and simulation techniques.
2. Celluar-level, or more recently, omics-level modelling, which examines all molecules of a particular kind (DNA, RNA, proteins, metabolites, *etc.*) of cells using high-throughput techniques.
3. Physiological-level, or system-level, which examines how drug and the body interact and affect each other.
4. Statistical modelling and inference, which underlies the three types of modelling above and informs us about the uncertainty of our models and inferences based on them.

## The process of drug discovery and development

A linear presentation of the drug discovery and development process can be often found in the following form or its variants.

```{r drugDiscoveryLinear}
```

The phases are briefly introduced below.

* Target assessment and validation
* Screening/target-to-hit
* Hit-to-lead
* Lead optimization
* Preclinical development
* Phase I clinical trial
* Phase II clinical trial
* Phase III clinical trial
* Submission to launch
* Post-launch activities

## Reflecting the linear model of drug discovery

The linear model of drug discovery is very useful in several aspects. First, it provides an initial mental model for people outside the industry and beginners to think about drug discovery. Second, it highlights the key decision points, in the chronological order, along the discovery and development process. Third, it simplifies communications about the stage in which drug candidates are currently in.

Despite the usefulness of the linear model, drug discovery in reality is rarely linear. Instead, it is often a dynamic and iterative process, with countless variants from the ideal linear model. During the discovery process, for instance, a lead compound may fail to show the expected efficacy in the preclinical experiment and therefore a new round of lead optimization is necessary or even a new hit series has to be identified. In another example, a drug that has been designed for one indication is found, either by real-world evidence or by serendipity, to be potentially effective for another indication, and thus the step of target identification, hit and lead optimization may be omitted or strongly shortened to meet the purpose. And in the development phase, more and more clinical trials are designed differently compared with the canonical ones, for instance with adaptive design, and with real-world data supported virtual trials. Many more examples like this suggest that the linear model is far away from the reality.

In the dynamic and iterative process, modelling mechanism and safety at different levels become important, because these models embody our belief of how drug works. Rationalized approaches may not always lead to successful drug-discovery projects; and very often. serendipity plays a critical role, especially in the discovery of drugs of new, unexpected mechanisms. 

Why, then, do we still care about rationalized approaches as described above in the multiscale modelling problem? First, because the rationalized approach complements the trial-and-error model of drug discovery. In addition, it can even pave ways to findings by chance, as long as the multiscale modelling process is not seen as a doctrine and the only way to success. Last but not least, the rationalized approach is the scientific way to follow up lucky findings in order to gain insights so that the drug candidate obtained by a lucky strike can be further improved in its efficacy and safety profiles. All the three aspects have been demonstrated by numerous examples in the last three decades of drug discovery.

## How will we approach the multiscale modelling problem in this course?

Generally, each session will follow a similar structure. We first list selected questions that can be potentially solved by mathematics and informatics tools, in the context of each stage in drug discovery. The questions are not comprehensive, but it is hoped that they give a flavour of real-world problems in the drug-discovery process. 

Next, we discuss models and concepts behind the tools that we choose, as well as the software tools that are used, in particular those open-source solutions. The list of models and concepts, similar like questions, are not comprehensive. In addition, they may be frequently used in practice but potentially suboptimal. The audience is invited to view them with a critical mind and challenge the routines and practices.

Finally, we will point to resources for further reading. This can contain two types of material, one type heavier on mathematics and informatics, and other type heavier on practice and process of drug discovery. My personal wish is that the audience can assess the resources that are more familiar to her or his own background, but also the resources on the other end of the spectrum. This is because, as we stated at the beginning of the chapter, drug discovery is an interdisciplinary enterprise, and this nature determines that interdisciplinary communication is key. Without knowing, using or mastering the language of other domains, we run in the risk of creating solutions that are not applicable to real problems.

## Applied mathematics and informatics in drug discovery

Applied mathematics and informatics have supported and transformed drug discovery from its early days to now. And the impact has been ever increasing thanks to the exponential growth of data, the fast progress of technology, and the ever more complicated diseases that we as human beings are facing with. 


Applications of mathematics and informatics can be found broadly in four categories:

### Molecular-level modelling

Molecular-level modelling of biological and chemical systems. This includes for instance biological sequence analysis and structural biology, which studies how biological sequences and variations are translated in structures and functions. The information are highly relevant for drug discovery, which aims at delaying, halting, or reversing disease progression by modulating biological molecules. On the the same level, physical models of both biological and chemical molecules are often used to study their interactions. Such chemical-biology interactions are fundamental to both efficacy and safety of drugs.

### Ceullar- and omics level modelling

What does a drug candidate do to cancer cells? Why does a compound with a particular structure cause DILI/toxicity in liver while another compound with a very similar structure does not? Such questions can be answered by cellular-level modelling. A human cell contains billions molecules, therefore it is not feasible to fully characterize the status of the whole system. However, it is possible to examine some aspects of the system in order to make inferences. For instance, in the second lecture, we will talk about the central dogma in molecular biology, which states that DNA directs transcription of RNA, which can be translated into proteins. While DNA is largely identical (not always so though) between different cells, the abundance of RNA and protein vary a lot depending on the context, for instance where the cells are, their interactions with other cells and external stimuli, what physiological status they are in (for instance activated or dying). Thus, it is feasible to measure all RNA or proteins in a cell (or the average of many cells) in order to query the status of the system. The techniques that we use to characterize all molecules of a given type is commonly known as the 'omics', which is a suffix coming from the Greek language.

Besides omics technologies, it is also possible to query other aspects of the cellular system. For instance, different types of imaging techniques such as microscopy and immunohistochemistry staining can be applied to directly observe whether treatment with an drug candidate cause morphological changes in cells. Combined with cell-type specific markers, it is possible to infer cell-type specific changes, which can be useful if the efficacy or toxicity of the drug is manifested in specific and/or rare cell types.

Apparently, cellular and omics level modelling misses the comprehensive and often deterministic nature of molecular modelling. And models on this level are of limited scope per definition. No change in RNA expression profiles of the cells, for instance, does not necessarily mean that the drug does not work (assuming that the drug reached the target and binds to it). Instead, for instance, the regulation may be only visible on the protein level. Therefore, change in omics profiles is neither necessary or sufficient for mechanisms of drugs. The same logic applies to imaging and other cellular and omics level models as well.

On the other side, cellular and omics level models bring its own strength and power. Molecular-level modelling often addresses one or one pair of molecules, for instance drug or drug-target protein interaction, in isolation. Due to limitations in our modelling power, we can hardly model more than a few molecules once a time. That requires we ignore many molecules in a cell that may interact with the system that we study. The ignorance has a price, which can manifested in unexpected off-target, nameling being active on another gene besides the one aimed at, effects of the drug. Even more, since proteins form networks to execute functions, even a highly specific modulator of a given protein will likely induce widespread consequences due to the fact that its target interacts with other proteins in the cell. This kind of signal propogation can hardly be studied using molucular levels, however is feasible to be studied using omics and cellular levels, when the prior knowledge of protein-protein interaction is integrated.

### 




## Some examples of bookdown

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@drawnel_molphen_2017].
