
```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.height=6, fig.width=6)
library(ribiosUtils)
library(ribiosIO)
library(ribiosPlot)
library(openxlsx)
library(tidyverse)
library(ggplot2)
library(emmeans)
library(knitr)
library(rmarkdown)
library(kableExtra)
theme_set(theme_light(base_size=14))
```

# Drug-target interaction

## Goals

* Drug-target interaction
* Binding assay versus functional assay
* Understand how Michaelis-Menten kinetics and dose-response curve are used to
  characterize drug-target interactions.

## Reversible binding of a drug to its target at a single site

The law of mass action assumes that the rate of a chemical reaction is
proportional to the product of the concentrations of the entities participating
the reaction, known as reactants. The law of mass action can be deduced from the collision theory.  The plot below gives an intuition

```{r collision, fig.height=5, fig.width=5}
library(spatstat.geom)
set.seed(1)
runif01 <- function(x) runif(x, 0, 1)
runif01LR <- function(nl, nr, scenario="Default scenario",
                      distCutoff=0.01) {
  lpoints <- spatstat.geom::runifrect(nl)
  rpoints <- spatstat.geom::runifrect(nr)
  df <- data.frame(scenario=scenario,
                    species=rep(c("Ligand", "Receptor"), c(nl, nr)),
                 x=c(lpoints$x, rpoints$x),
                 y=c(lpoints$y, rpoints$y))
  lrdist <- spatstat.geom::crossdist(lpoints, rpoints)
  whichIsClose <- which(lrdist <= distCutoff, arr.ind = TRUE)
  lcloseInd <- whichIsClose[,1]
  rcloseInd <- whichIsClose[,2]
  xclose <- (lpoints$x[lcloseInd] + rpoints$x[rcloseInd])/2
  yclose <- (lpoints$y[lcloseInd] + rpoints$y[rcloseInd])/2
  closeMidpoints <- data.frame(scenario=scenario, x=xclose, y=yclose, radius=distCutoff)
  res <- list(df=df, collisions=closeMidpoints)
  return(res)
}
circleFun <- function(center = matrix(c(0,0), nrow=1, by.row=TRUE),
                      group=1:nrow(center),
                      radius = 1, npoints = 100){
    tt <- seq(0,2*pi,length.out = npoints)
    groupOut <- rep(group, each=length(tt))
    point <- rep(1:nrow(center), each=length(tt))
    xx <- rep(center[,1], each=length(tt)) + radius * cos(tt)
    yy <- rep(center[,2], each=length(tt)) + radius * sin(tt)
    return(data.frame(group=groupOut, pointInd=point, x = xx, y = yy))
}
s1 <- runif01LR(50, 50, "Scenario 1")
s2 <- runif01LR(500, 50, "Scenario 2")
s3 <- runif01LR(50, 500, "Scenario 3")
s4 <- runif01LR(500, 500, "Scenario 4")
LRscenarios <- rbind(s1$df, s2$df, s3$df, s4$df)
collisions <- rbind(s1$collisions, s2$collisions, s3$collisions, s4$collisions)
collisionCircles <- circleFun(center=cbind(collisions$x,
                                           collisions$y), 
                              group=collisions$scenario,
                              radius=collisions$radius[1]) %>%
  mutate(scenario=group, color="Collision") %>%
  mutate(group_point=interaction(group, pointInd, drop = TRUE))
ggplot(LRscenarios, aes(x=x, y=y, col=species)) +
  facet_wrap(~scenario) +
  geom_point(pch=16, size=1) +
  scale_color_manual(values=c("Ligand"="#1f78b4", "Receptor"="#b2df8a",
                              "Collision"="#e41a1c")) +
  xlab("") + ylab("") +
  geom_path(data=collisionCircles, aes(x=x, y=y, group=group_point, col=color), 
            inherit.aes = FALSE, lwd=1,
            show.legend=FALSE) +
  scale_fill_manual(name="Colission",values=c("Collision"="#e41a1c")) +
  guides(colour = guide_legend(override.aes = list(size=3), title="")) +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "top",
        legend.margin = margin(), legend.box.margin = margin(),
        panel.border = element_rect(fill="transparent", colour="#333333"))
```


$$
\require{mhchem}
\ce{L + R <=>[k_1][k_2] LR}
$$

Consider the situation where a drug (ligand) binds reversibly to a target
(a receptor, or more generally a protein) at a single site. We can write the following ordinary differential
equation with regard to the production rate of the complex $[LR]$.

$$ \frac{d[LR]}{dt} = k_1[L][R] - k_2[LR] $$.

When the reaction reaches its equilibrium, namely when the production rate is
constant at zero, we have

$$ k_1[L][R] = k_2[LR] $$.

Note that $R_{total}=[R] + [LR]$, so we have

\begin{align}
k_1[L]([R_{total}]-[LR]) & = k_2[LR] , \\
[LR] &= \frac{k_1[L][R_{total}]}{k_1[L]+k_2}.
\end{align}

If we define $K_D \equiv k_2/k_1$, we have

$$ [LR] = [R_{total}]\frac{[L]}{[L]+K_D} $$

If we plot the relationship between $[LR]$ and $[L]$, we observe a rectangular
hyperbola curve.

```{r hyperbola, fig.height=4, fig.width=4}
kd <- 35
l <- 1:1000
r_total <- 1
lr <- r_total * (l)/(l+kd)
hyperbola_df <- data.frame(l=l,
                           lr=lr)
ggplot(hyperbola_df, aes(x=l, y=lr)) +
  geom_line() +
  geom_vline(xintercept = kd, lty=3, show.legend = TRUE) +
  geom_hline(yintercept = 1, lty=2) +
  scale_linetype_manual(name="", values=c("Full occupancy"=2,
                                          "KD"=3)) +
  annotate("text", x=quantile(l, 0.8), y=1.02, 
           label="Full occupancy", vjust=0) +
  annotate("text", x=kd*1.2, y=0.5, vjust=0, hjust=0,
           label= "K[D]", parse=TRUE) +
  scale_y_continuous(labels=scales::percent,
                     breaks = seq(0,1, 0.25),
                     expand=expansion(add=0.10)) +
  xlab("Concentration of L (arbitrary unit)") + ylab("Relative occupancy of R")
```
```{r hyperbola_log, fig.height=4, fig.width=4}
ggplot(hyperbola_df, aes(x=l, y=lr)) +
  geom_line() +
  geom_vline(xintercept = kd, lty=3, show.legend = TRUE) +
  geom_hline(yintercept = 1, lty=2) +
  scale_linetype_manual(name="", values=c("Full occupancy"=2,
                                          "KD"=3)) +
  annotate("text", x=quantile(l, 0.3), y=1.02, 
           label="Full occupancy", vjust=0) +
  annotate("text", x=kd*1.2, y=0.5, vjust=0, hjust=0,
           label= "K[D]", parse=TRUE) +
  scale_x_log10() +
  xlab("concentration of L (log10(arbitrary unit))") + ylab("Relative occupancy of R")
```

We can interpret $K_D$ in two ways. First, it is the ratio of the velocity of
the reverse reaction over that of the forward reaction. The smaller the value
is, the more quickly the ligand-receptor target is formed. Second, notice that
when $[L]=K_D$, $[LR]=[R_{total}]/2$. $K_D$ there represents the concentration
of ligand that occupies 50 percent of all receptors.

From physically chemistry we learn that thermodynamics studies can determine the
relative contributions of enthalpy and entropy of a compound's binding energy.
Roughly speaking, the enthalpy contribution describes whether a chemical
reaction absorbs or releases energy from or to the environment, and the entropy
change measures whether the degree of randomness or disorder of the system
increases or decreases.

The Gibbs free energy of binding ($\Delta G$) is made up of enthalpic and
entropic contributions. For reversible binding events, it is described as

$$ \Delta G = \Delta H-T\Delta S $$

The Van't Hoff equation defines the relationship between binding affinity,
$K_D$, and the temperature, $T$:

$$ ln(K_D) = \frac{\Delta H}{R}\frac{1}{T}-\frac{\Delta S}{R}, $$

where $R$ is the Gas Constant. By probing binding kinetics at varying
temperatures, and performing linear regression analysis of $K_D$ versus $1/T$,
we can infer the relative contribution of enthalpy and entropy to the compound's
binding energy. The information is valuable because chemists can then optimize
either enthalpy or entropy contribution or both by modifying chemical
structures. In drug design, it is desired to increase enthalpic contribution by
optimising ligand-target interactions , for instance by introducing
electrostatic interactions, hydrogen bonding, *etc.*. At the same time, it is
desired to decrease entropic (conformational) contributions by making the compound more rigid, for instance limiting the rotational freedom on binding.

Therefore, we can measure the direct binding of a compound to a target by
varying its concentration and measuring the concentration of ligand-protein
complex. The dissociation constant $K_D$ can be estimated from the midpoint of
the saturation curve. This however requires that we can measure the
concentration of ligand-receptor complex, which in turn requiring us label the
test compound, which is not usually feasible.

Alternatively, one can test the competitive binding of a test compound and a
labelled compound by displacing the labelled compound with increasing
concentrations of a test compound. This provides an indirect assessment of
binding.

## The Michaelis-Menten kinetics

The Michaelis-Menten kinetics model, named after German biochemist Leonor
Michaelis and Canadian physician Maud Menten, describes the relationship between
the rate of enzymatic reactions and the rate of formation of the product as well
as the concentration of the substrate. It is often used to model biochemical
reactions involving a single substrate.

The Michaelis-Menten kinetics have the following form:

$$ v = \frac{d[P]}{dt} = \frac{V_{max}[S]}{K_{Ḿ}+[S]}, $$

where,

* $V_{max}$ represents the maximum rate achieved by the system, which is reached
  at the saturating substrate concentration,
* $K_{M}$ is known as the Michaelis constant, numerically equal to the substrate
  concentration at which the reaction rate is half of $V_{max}$.

Given a enzyme reaction described below, we can derive the Michaelis-Menten
model by starting from the law of mass action.

$$
\require{mhchem}
\ce{E + S <=>[k_f][k_r] ES ->[k_{cat}] E + P}
$$

In the reaction above, the law of mass action will entitle four non-linear
ordinary differential equations that define the rate of change of reactants,
with regard to time $t$,

\begin{align}
\frac{d[E]}{dt} &= -k_f[E][S] + k_r[ES]+k_{cat}[ES], \\
\frac{d[S]}{dt} &= -k_f[E][S] + k_r[ES], \\
\frac{d[ES]}{dt} &= k_f[E][S] - k_r[ES] - k_{cat}[ES], \\
\frac{d[P]}{dt} &= k_{cat}[ES],
\end{align}

where

* $[E]$, $[S]$, $[ES]$ and $[P]$ represent concentration of the enzyme, the
  substrate, the enzyme-substrate complex, and the product, respectively.
  Because the enzyme $E$ is only a catalyst, its total concentration remain
  constant, namely $[E]+[ES]=[E]_{total}$. We can also write $[E]_{total}$ as
  $[E]_0$.
* $k_f$, $k_r$, and $k_{cat}$ represent the rate constant of the forward
  reaction, the reverse reaction, and the catalytic reaction, respectively.

While it is often feasible (though not always easy) to determine $[E]$, $[S]$,
and $[P]$ analytically, because their concentration is either specified or
easily measurable, the quantification of $[ES]$ is usually more challenging,
because the complex usually exists transiently in a low amount. This fact,
however, suggests that we can simplify the ODE system by making certain
assumptions about the chemical system that we study. For instance, we may assume
that the substrate is in *instantaneous* chemical equilibrium with the complex,
which suggests

$$ k_{f}[E][S] = k_{r}[ES]. $$

Based on the fact that $[E] = [E]_0 - [ES]$, we can deduce the Michaelis-Menten
model in the following way,

\begin{align}
k_f([E]_0 - [ES])[S] &= k_{r}[ES] \\
k_f[E]_0[S]-k_f[ES][S] &= k_{r}[ES] \\
k_f[E]_0[S] &= k_{r}[ES]+k_f[ES][S] \\
k_f[E]_0[S] &= [ES](k_{r}+k_f[S]) \\
[ES] &= \frac{k_f[E]_0[S]}{k_r + k_f[S]} \\
[ES] &= \frac{k_f[E]_0[S]}{k_f(\frac{k_r}{k_f} + [S])} \\
\end{align}

If we define the dissociation constant $K_D \equiv \frac{k_r}{k_f}$, we get

$$ [ES] = \frac{[E]_0[S]}{K_D+[S]} $$

Therefore, the velocity of the reaction, $v$, defined as the rate at which the
product $P$ is formed, can be expressed as

$$ v = \frac{d[P]}{dt} = k_{cat}[ES] = \frac{k_{cat}[E]_0[S]}{K_D + [S]} $$

Assuming that at a certain time point in a reaction, all enzymes form $ES$
complexes, the velocity of the reaction should then reaches its maximum, because $max([ES])=[E]_0$, when $[E]=0$. We can therefore define $V_{max} \equiv k_{cat}[E]_0$. Therefore

$$ v = \frac{V_{max}[S]}{K_D + [S]} $$

An alternative assumption known as the quasi-steady-state assumption of the
pseudo-steady-state assumption, proposed by Briggs and Haldane, is that the
concentration of the intermediate complex remains constant on the time scale of
product formation. It can be expressed mathematically as $k_f[E][S] =
k_r[ES]+k_{cat}[ES] = (k_r+k_{cat})[ES]$. Apparently, this has essentially the
same mathematical formulation as the assumption above, with $k_r$ replaced by
$k_r+k_{cat}$. If we accept this assumption, the form of the Michealis-Menten
model remains largely unchanged, with the only exception that $K_D$ is replaced
by $K_M$, the Michaelis constant, defined as $K_M \equiv \frac{k_r + k_{cat}}{k_f}
$. Whereas the first assumption is valid if the substrate reaches equilibrium on
a much faster time-scale than the product is formed, the second assumption hold
true if the enzyme concentration is much less than $[S]$ or $K_M$ or both.

In practice, to determine $V_{max}$ and $K_M$, a series of enzyme assays can be
run with varying substrate concentrations $[S]$ and measuring the initial
reaction rate $v_0$. 

## Agonist dose-response curves

In functional assays, we often use the Hill equation, a four- or five-parameter
function which has a sigmoidal curve, to describe the consequence of binding of
ligand to a target. The Hill equation has two forms: the Hill-Langmuir equation
is derived from the Langmuir absorption model; the Hill equation reflects the
cellular and tissue response to a ligand. In either case, geometrically, the
Hill equation is a special case of a rectangular hyperbola, and has the general
form of

$$ H = H_{max}\frac{x^n}{k^n+x^n} $$

For instance, the response of a cell (for instance, apoptosis, or programmed
cell death) to a drug candidate may be modelled by the following equation

\begin{align}
E &= E_{max} \frac{[L]^n}{EC_{50}^n + [L]^n} \\
  &= E_{max} \frac{1}{1+(\frac{EC_{50}}{[L]})^n}
\end{align}

where

* $[L]$ is the ligand concentration,
* $[EC_{50}]$ is the drug concentration that produces a 50% maximal response.

The Hill equation represents an example of dose-response curve. Such curves are
critical for drug candidate characterization and characterization.

## Deterministic and stochastic simulations

### Deterministic simulation

See http://seweb.se.wtb.tue.nl/biological_systems/de.

$$
\require{mhchem}
\ce{S + E <=>[k_1][k_2] C ->[k_3] P + E }
$$

And according to the law of mass action

\begin{align}
\frac{d[S]}{dt} &= -k_1[E][S] + k_2[C], \\
\frac{d[E]}{dt} &= -k_1[E][S] + (k_{2}+k_{3})[C], \\
\frac{d[C]}{dt} &= k_1[E][S] - (k_{2}+k_{3})[C], \\
\frac{d[P]}{dt} &= k_{3}[C],
\end{align}

If the initial concentrations and the reaction rates are given, the set of ODEs
can be numerically solved. A deterministic simulation therefore is possible. For
instance with the following examples (*Stochastic Modelling for Systems
Biology*).

* $S(0)=5e^{-7}$
* $E(0)=2e^{-7}$
* $C(0)=P(0)=0$
* $k_1=1e^6$
* $k_2=1e^{-4}$
* $k_3=0.1$

### Stochastic modelling

Good reference: Higham, Desmond J. 2008. “Modeling and Simulating Chemical Reactions.” SIAM Review 50 (2): 347–68. https://doi.org/10.1137/060666457.

We can formulate a chemical reaction with the *chemical master equation* (CME),
a set of ordinary differential equations (ODEs), each ODE representing one
possible state of the system. At time $t$, the $k$th equation gives the
probability of the system in the $k$th state. In this case, the dimension of the
ODE is not given by the number of species $N$, but by the number of possible
states of the system. It depends on the total number of molecules as well as the
form of the chemical reaction.

Suppose we have following reactions

$$
\require{mhchem}
\ce{A + B <=>[k_1][k_2] C + D }
$$

If we start with $K$ molecules of type A and $K$ molecules of type B, and no
molecules of either $C$ or $D$, then we have

$$
\begin{align}
    X(0) &= \begin{bmatrix}
           K \\
           K \\
           0 \\
           0
         \end{bmatrix},
  \end{align}
$$

then the state vector $X(t)$ takes one of the following possible values

$$
\begin{bmatrix}
           K \\
           K \\
           0 \\
           0
         \end{bmatrix},
\begin{bmatrix}
           K-1 \\
           K-1 \\
           1 \\
           1
         \end{bmatrix},
\begin{bmatrix}
           K-2 \\
           K-2 \\
           2 \\
           2
         \end{bmatrix},
\ldots,
\begin{bmatrix}
           0 \\
           0 \\
           K \\
           K
         \end{bmatrix},
$$
