# Chapter eighteen: population modelling

These resources can be interesting

1. Open Systems Pharmacology Community 
2. See comments here: https://www.researchgate.net/post/Is_there_any_open_source_Pharmacokinetic_software
3. FDA RWE and RWD guideline: https://www.fda.gov/science-research/science-and-research-special-topics/real-world-evidence

## The nonlinear mixed effect model

Let us start with the linear case

```{r, fig.height=4, fig.width=4}
library(rethinking)
library(dplyr)
library(ggplot2)
data(Howell1)
d2 <- Howell1 %>% filter(age>=18) %>%
  mutate(weight_s=scale(weight))
d2plot <- ggplot(d2, aes(x=weight_s, y=height)) +
  geom_point() +
  ggtitle("The !Kung San data (adults only)") +
  xlab("Standarized weight")
print(d2plot)
```

Using frequentist's language

$$
  y_{i} = f(x) + \epsilon \\
  f(x) = \alpha + \beta x \\
  \epsilon \sim \mathcal{N}(0, \sigma)
$$
Using Bayesian language and some personal prior

$$
  y_{i} \sim \mathcal{N}(\mu_{i}, \sigma) \\
  \mu = \alpha + \beta x_{i} \\
  \alpha \sim \mathcal{N}(169, 20) \\
  \sigma \sim \textrm{Unfoirm}(0, 50)
$$

```{r, fig.height=4, fig.width=4}
d2plot + stat_smooth(method='lm', formula= y~x)
```

Consider a one-compartment model with drug administered as a single intravenous bolus dose

$$
C(t) = \frac{\textrm{Dose}}{V}e^{-\frac{CL}{V}t}
$$

$$
  y_{ij} \sim \mathcal{N}(\mu_{ij}, \Sigma_{i}) \\
  \mu_{ij} = f(t_{ij}, \beta_{i}, d_{i}) \\
  \beta_{i} \sim \mathcal{N}(\beta, D)
$$
Parameters:

* $y_{ij}$ is the $j^{th}$ response for the $i^{th}$ subject
* $f$ is a scalar function nonlinear with regard to $\beta$
* $\beta$ is a $k \times 1$ parameter vector, giving PK parameters such as absorption, $V_D$, and $CL$.
* $t_{ij}$ is the $j^{th}$ time of measurement for the $i^{th}$ subject
* $d_{i}$ is the dose of the $i^{th}$ subject
* $j$ ranges from 1 to $n_i$
* $D$ is a $k \times k$ covariance matrix
* $\Sigma_i$ is an $n_{i} \times n_{i}$ covariance matrix
